Efficient stability for the Pauli group

### 2023-06-05 efficient_stability.tex
continuing the proof of 5.1

To describe the tester we need to introduce interpolation coefficients, as follows. For $u\in \F_q$ and $i\in\{1,\ldots,d\}$ let 
\[ \alpha_{u,i} = \frac{\prod_{\substack{i'=1\\i'\neq i}}^{d+1} i'}{\prod_{\substack{i'=1\\i'\neq i}}^{d+1} (i'-i)}\;.\]
These are defined so that for any $f:\F_q\to\F_q$ of degree at most $d$, $f(u)=\sum_{i=1}^{d+1} \alpha_{u,i} f(u+i)$. The tester verifies this relation along a randomly chosen direction. 
\tnote{Here I chose to specialize the ``line'' to $(d+2)$ consecutive points on it, which is notationally marginally simpler than $(d+2)$ uniformly random points on it. Hopefully it's fine}

@@

\begin{claim}\label{claim:z2-stab-2}
Let $q=2^t$. 
For every $u\in \F_q^m$ there is a projective measurement $\{P^u_\alpha\}_{\alpha\in \F_q}$ on $\mM$ such that 
\[ \Es{u\in \F_q^m} \Es{\alpha\in \F_2^t} \big\| \phi(s_{u,\alpha}) - \sum_{\beta\in\F_2^t} (-1)^{\alpha \cdot \beta} P^u_\beta \big\|_\tau^2 \,=\, O(\eps)\;. \]
\end{claim}

@@

The following claim will prove useful.

\begin{claim}\label{claim:z2-stab-2b}\tnote{This will require a specific distribution on the commutation relations}
\[ \Es{u\in \F_q^m} \Es{j\in\{1,\ldots,m\}} \Es{i\neq i' \in \{0,\ldots,d+1\}} \big\| [P^{u+ie_j}_\alpha, P^{u+i'
 e_{j}}_\beta ]\big\|_\tau^2 = O(\eps)\;. \]
\end{claim}

\begin{proof}
From Claim~\ref{claim:z2-stab-2} we have that 
\[ \Es{u\in \F_q^m} \Es{\alpha\in \F_2^t} \big\|\Es{\alpha\in \F_2^t} (-1)^{\alpha \cdot \beta} \phi(s_{u,\alpha}) -  P^u_\beta \big\|_\tau^2 \,=\, O(\eps)\;. \]
The claim therefore follows from the commutation relations. 
\end{proof}

@@

\begin{claim}\label{claim:z2-stab-3}
For every axis-parallel line $\ell\subseteq \F_q^m$ there is a projective measurement $\{Q^\ell_g\}$ on $\mM$ such that 
\[ \Es{u\in \F_q^m} \sum_f \tau\big( P^u_{f(u)} Q^\ell_g\big) \,\geq\, 1-\poly(\eps)\;. \]
\end{claim}

\begin{proof}
Fix a direction $j\in \{1,\ldots,m\}$, an $u\in \F_q^m$, and let $u_i = u+ i e_j$ for $i\in \{0,\ldots,d+1\}$. Let $\{P^{u_i}_\alpha\}$ be the projective measurement obtained from Claim~\ref{claim:z2-stab-2}, and $U_i = \sum_\alpha \omega_q^\alpha P^u_\alpha$, where $\omega_q = e^{2i\pi/q}$. Then $U_i \in \mU(\mM)$. Furthermore, from Claim~\ref{claim:z2-stab-2b} we immediately get that
\begin{equation}\label{eq:z2-stab-3-1}
\Es{i\neq j \in \{0,\ldots,d+1\}} \big\| [U_i,U_j] \big\|_\tau^2 = O(\eps)\;.
\end{equation}
We define $\psi: \Z_q^{d+2} \to \mU(\mM)$ by 
\[\psi(x_1,\ldots,x_{d+2}) \,=\, U_1^{x_1} \cdots U_{d+1}^{x_{d+1}}\;,\]
which is well-defined since each $U_i$ has order $q$. Using~\eqref{eq:z2-stab-3-1}, the map $\psi$ is an $O(d\eps)$-approximate homomorphism of $\Z_q^{d+2}$ on $\mU(\mA)$.

This allows us to apply Theorem~\ref{thm:gh} to obtain a partial isometry $w\in P\mM_\infty\Id_\mM$ and a family of commuting unitaries $\{V_i\}$ of order $q$ each on $\mN=P\mM_\infty P$ such that 
\[ \Es{i\in \{0,\ldots,d+1\}} \big\| U_i - w^* V_i \big\|_\tau^2 \,=\, O(d\eps)\;.\]
\tnote{Actually this is not quite the right distance measure}
Because $\{V_i\}$ are a representation of $\F_q^{d+1}$, there is a projective measurement $\{Q_\alpha\}_{\alpha\in \F_q^{d+1}}$ on $\mN$ such that $U_\alpha = \sum_\beta (-1)^{\alpha\cdot \beta} Q^u_\beta$. ($\{Q^u_\alpha\}$ can be found explicitly by applying the Fourier transform.) Applying Lemma~\ref{lem:pull-back}, we deduce a projective measurement $\{P^u_\alpha\}_{\alpha\in \F_2^t}$  on $\mM$ such that (by the triangle inequality)
\[ \Es{\alpha \in \F_2^t} \big\| \phi(s_{u,\alpha}) - \sum_{\beta\in\F_2^t} (-1)^{\alpha \cdot \beta} P^u_\beta \big\|_\tau^2 \,=\, O(\eps_u)\;.\]

Let $V_1,\ldots,V_{d+1}$ be commuting unitaries. POVM. Pull-back.

>>>>

To describe the tester we need to introduce interpolation coefficients, as follows. For $u\in \F_q$ and $i\in\{1,\ldots,d\}$ let 
\[ \alpha_{u,i} = \frac{\prod_{\substack{i'=1\\i'\neq i}}^{d+1} i'}{\prod_{\substack{i'=1\\i'\neq i}}^{d+1} (i'-i)}\;.\]
These are defined so that for any $f:\F_q\to\F_q$ of degree at most $d$, $f(u)=\sum_{i=1}^{d+1} \alpha_{u,i} f(u+i)$. The tester verifies this relation along a randomly chosen direction. 
\tnote{Here I chose to specialize the ``line'' to $(d+2)$ consecutive points on it, which is notationally marginally simpler than $(d+2)$ uniformly random points on it. Hopefully it's fine}\tnote{edit: I think I want to have the line be $(d+1)$ consecutive points, and also query at another independent random point on the same line to check the interpolation}

@@

\begin{claim}\label{claim:z2-stab-2}
Let $q=2^t$. 
For every $u\in \F_q^m$ there is a projective measurement $\{P^u_\alpha\}_{\alpha\in \F_{2^t}}$ on $\mM$ such that 
\begin{equation}\label{eq:z2-stab-2}
 \Es{u\in \F_q^m} \Es{\alpha\in \F_2^t} \big\| \phi(s_{u,\alpha}) - \sum_{\beta\in\F_2^t} (-1)^{\alpha \cdot \beta} P^u_\beta \big\|_\tau^2 \,=\, O(\eps)\;. 
\end{equation}
\end{claim}

@@

For $u\in \F_q^m$ let $\{P^{u}_\alpha\}_{\alpha \in \F_2^t}$ be the projective measurement obtained from Claim~\ref{claim:z2-stab-2}, and 
\[ U_u = \sum_{\alpha\in\F_q} \omega_q^\alpha P^u_{\kappa(\alpha)}\;,\]
where $\omega_q = e^{2i\pi/q}$. Then $U_i \in \mU(\mM)$.

%We observe that
%\begin{equation}\label{eq:z2-stab-2a}
%\Es{u\in\F_q^m}\Es{\alpha\in \F_q} \big\| U_u^\alpha - \phi(s_{u,\kappa(\alpha)})\big\|_\tau^2 \,=\, O(\eps)\;.
%\end{equation}
%To see this, observe that for any $\alpha\in \F_q$, $\omega_q^\alpha = (-1)^{\tr(\alpha)}$. Thus from the definition of $U_u$ we get
%\begin{align*}
%U_u^\alpha &= \sum_\beta \omega_q^{\alpha\beta} P^u_\alpha\\
 %&= \sum_\beta (-1)^{\tr(\alpha\beta)} P^u_\alpha\\
%&= \sum_\beta (-1)^{\alpha\cdot \beta} P^u_\alpha\;,
%\end{align*}
%where in the last line we slightly abused notation and identified $\alpha,\beta\in\F_q$ with $\kappa(\alpha),\kappa(\beta)\in \F_2^t$ respectively. \eqref{eq:z2-stab-2a} then follows from~\eqref{eq:z2-stab-2}.

\begin{claim}\label{claim:z2-stab-2b}\tnote{This will require a specific distribution on the commutation relations}
For $u\in \F_q^m$, $j\in\{1,\ldots,m\}$ and $i\in \{0,\ldots,d+1\}$ let $U_{i} = U_{u+ie_j}$ and $P^{(i)}_\alpha = P^{u+ie_j}_\alpha$ (leaving the dependence on $u$ implicit). Then 
\[ \Es{u\in \F_q^m} \Es{\substack{j\in\{1,\ldots,m\}\\i\neq i' \in \{0,\ldots,d+1\}}}\Es{\alpha,\alpha'\in \F_q} \big\| \big[ U_i^\alpha, U_{i'}^{\alpha'}\big]-\Id\big\|_\tau^2\,=\, O\big(\sqrt{\eps}\big)\;. \]
\end{claim}

\begin{proof}
Due to the test of the relations $\{R^\com_k\}$, it holds that \tnote{this will depend on the distribution placed on commutation relations} 
\begin{equation}\label{eq:z2-stab-2b-1}
\Es{u\in \F_q^m} \Es{\substack{j\in\{1,\ldots,m\} \\ i\neq i' \in \{0,\ldots,d+1\}}} \Es{\alpha,\beta\in \F_q} \big\| [\phi(s_{u+i e_j,\alpha}),\phi(s_{u+i' e_j,\beta})] - \Id \big\|_\tau^2 = O(\eps)\; .
\end{equation}
Now, for $u\in \F_q^m$, $j\in \{1,\ldots,m\}$ and $i,i'\in \{0,\ldots,d+1\}$,
\begin{align*}
\Es{\alpha,\alpha'\in \F_q} &\tau\big( U_i^\alpha U_{i'}^{\alpha'} U_i^{-\alpha} U_{i'}^{-\alpha'} \big) \\
&= \sum_{\beta,\beta',\gamma,\gamma'\in \F_q}\Es{\alpha\in \F_q} \omega_q^{\alpha(\beta-\beta')} \Es{\alpha'\in \F_q} \omega_q^{\alpha'(\gamma-\gamma')} \tau(P^{(i)}_{\kappa(\beta)} P^{(i')}_{\kappa(\gamma)} P^{(i)}_{\kappa(\beta')} P^{(i')}_{\kappa(\gamma')}\big)\\
&=\sum_{\beta,\beta',\gamma,\gamma'\in \F_2^t}\Es{\alpha\in\F_2^t} (-1)^{\alpha(\beta-\beta')} \Es{\alpha'\in\F_2^t} (-1)^{\alpha'(\gamma-\gamma')} \tau(P^{(i)}_\beta P^{(i')}_\gamma P^{(i)}_{\beta'} P^{(i')}_{\gamma'}\big)\\
&=\Es{\alpha,\alpha'\in\F_2^t}  \tau\Big( \Big(\sum_{\beta\in \F_2^t} (-1)^{\alpha\cdot \beta} P^{(i)}_\beta\Big) \Big(\sum_{\gamma\in \F_2^t} (-1)^{\alpha'\cdot \gamma} P^{(i')}_\gamma\Big)\Big(\sum_{\beta'\in \F_2^t} (-1)^{\alpha\cdot \beta'} P^{(i)}_{\beta'}\Big)\Big(\sum_{\gamma'\in \F_2^t} (-1)^{\alpha'\cdot \gamma'} P^{(i')}_{\gamma'}\Big)\Big)\;.
\end{align*}
Here, the second line follows since for $\beta,\beta'\in \F_q$,
\[ \Es{\alpha\in \F_q} \omega_q^{\alpha(\beta-\beta')}\,=\,\Es{\alpha\in\F_2^t} (-1)^{\alpha(\kappa(\beta)-\kappa(\beta'))}\,=\delta_{\beta,\beta'}\;.\]
It follows that, on average over uniformly random $u\in \F_q^m$, $j\in \{1,\ldots,m\}$ and $i,i'\in \{0,\ldots,d+1\}$,
\begin{align*}
\Es{\alpha,\alpha'\in \F_q} \big\| [U_i^\alpha ,U_{i'}^{\alpha'} ]-\Id\big\|_\tau^2
&= \Es{\alpha,\alpha'\in \F_2^t} \Big\| \Big[\sum_{\beta\in \F_2^t} (-1)^{\alpha\cdot \beta} P^{(i)}_\beta ,\sum_{\beta'\in \F_2^t} (-1)^{\alpha\cdot \beta'} P^{(i')}_{\beta'} \Big]-\Id\Big\|_\tau^2\\
&= \Es{\alpha,\alpha'\in \F_2^t} \big\| \big[ \phi(s_{u+i e_j,\alpha}),\phi(s_{u+i'e_j,\alpha'})\big]-\Id\big\|_\tau^2+O\big(\sqrt{\eps}\big)\;,
\end{align*}
where the second line follows from Claim~\ref{claim:z2-stab-2}. This last expression is bounded by~\eqref{eq:z2-stab-2b-1}.
\end{proof}

@@

\begin{claim}\label{claim:z2-stab-3}
For every $u\in \F_q^m$ and axis-parallel line $\ell\subseteq \F_q^m$ through $u$ in the $j$-th direction there is a projective measurement $\{Q^{u,j}_g\}$ on $\mM$ such that 
\[  \Es{x_0,\ldots,x_{d}\in \F_q} \Big\| U_0^{x_0}\cdots U_d^{x_d} -  \sum_{g} \omega^{\sum x_ig(u+ie_j)} R^{u,j}_g\Big\|_\tau^2 \,=\, \poly(d,\eps)\;.\]
\end{claim}

\begin{proof}
Fix an $u\in \F_q^m$ and a direction $j\in \{1,\ldots,m\}$. Define
$\psi: \Z_q^{d+1} \to \mU(\mM)$ by 
\[\psi(x_0,\ldots,x_{d}) \,=\, U_0^{x_0} \cdots U_{d}^{x_{d}}\;,\]
where for each $i\in\{0,\ldots,d\}$, $U_i$ is the order-$q$ unitary defined in Claim~\ref{claim:z2-stab-2b}. Using Claim~\ref{claim:z2-stab-2b} and the triangle inequality we verify that the map $\psi$ is a $\delta=O(d^2\sqrt{\eps})$-approximate homomorphism of $\Z_q^{d+1}$ on $\mU(\mA)$, with respect to the exhaustive presentation\tnote{need to give details and check exponents of $d$,$\eps$}.

Thus we may apply Theorem~\ref{thm:gh} to obtain a partial isometry $w\in P\mM_\infty\Id_\mM$ and a family of commuting unitaries $\{V_i\}$ of order $q$ each on $\mN=P\mM_\infty P$ such that, on average over $u$ and $j$,
\[ \Es{x_0,\ldots,x_{d}} \big\| \psi(x_0,\ldots,x_{d+1}) - w^* V_0^{x_0}\cdots V_{d}^{x_{d}} w \big\|_\tau^2 \,=\, \poly(d,\eps)\;.\]
Because $\{V_i\}$ are a representation of $\Z_q^{d+1}$ (which is Abelian), there is a projective measurement $\{Q_\alpha\}_{\alpha\in \F_q^{d+1}}$ on $\mN$ such that for each $i\in\{0,\ldots,d\}$, 
\[ V_i \,=\, \sum_{\alpha\in \Z_q^{d+1}} \omega^{\alpha_i} Q_\alpha\;,\]
and  
\[ V_0^{x_0}\cdots V_{d}^{x_{d}}  \,=\, \sum_{\alpha\in \Z_q^{d+1}} \omega^{\sum x_i\alpha_i} Q_\alpha\;.\]
%($\{Q^u_\alpha\}$ can be found explicitly by applying the Fourier transform.) 
Applying Lemma~\ref{lem:pull-back}, we deduce a projective measurement $\{R^u_\alpha\}_{\alpha\in \Z_q^{d+1}}$  on $\mM$ such that, on average over $u$ and $j$,
\[  \Es{x_0,\ldots,x_{d}} \Big\| \psi(x_0,\ldots,x_{d+1}) -  \sum_{\alpha\in \Z_q^{d+1}} \omega^{\sum x_i\alpha_i} R_\alpha\Big\|_\tau^2 \,=\, \poly(d,\eps)\;.\]
For $\ell$ the line going through $u$ and parallel to the $j$-th axis, define $Q^\ell_g = R_{(g(u),\ldots,g(u+de_i))}$.
\end{proof}

Next we show that the statement of Claim~\ref{claim:z2-stab-3} can be strengthened to \emph{every} $x_0,\ldots,x_{d}$, instead of on average. This is done in the next claim.

\begin{claim}\label{claim:z2-stab-4}
With the same notation as in Claim~\ref{claim:z2-stab-3}, for any $x_0,\ldots,x_d\in \F_q$ it holds that
\[  \Big\| U_0^{x_0}\cdots U_d^{x_d} -  \sum_{g} \omega^{\sum x_ig(u+ie_j)} R^{u,j}_g\Big\|_\tau^2 \,=\, \poly(d,\eps)\;.\]
\end{claim}

\begin{proof}
For any $y_0,\ldots,y_d\in \F_q$ we can write 
\[ U_0^{x_0}\cdots U_d^{x_d}\,=\, U_0^{y_0}U_0^{x_0-y_0}\cdots U_d^{y_d}U_0^{x_0-y_0}\;.\]
By repeated application of Claim~\ref{claim:z2-stab-2b}, the term on the right-hand side is 
\[  U_0^{y_0}U_0^{x_0-y_0}\cdots U_d^{y_d}U_0^{x_0-y_0} \,\simeq_{\poly(d,\eps)}U_0^{y_0}\cdots U_d^{y_d}U_0^{x_0-y_0}\cdots U_d^{x_d-y_d}\;.\]
To show this it suffices to verify that we only need to ``commute'' pairs of terms whose exponents are indepent and uniformly random. The conclusion then follows by two applications of Claim~\ref{claim:z2-stab-3}.
\end{proof}

Applying Claim~\ref{claim:z2-stab-4} to the interpolation coefficients $x_0,\ldots,x_d$, by Fourier transform we obtain
\[ \Es{u\in \F_q^m} \sum_g \tau\big( P^u_{g(u)} Q^\ell_g\big) \,\geq\, 1-\poly(\eps)\;. \]
\tnote{This should be one more claim. There are some parts to handle, e.g. the distribution on points, etc.}

\tnote{Now we are ready to define a strategy and check that it succeeds in the tensor-codes game}
